TrenchR: an R package for transparent environmental and ecological biophysics

### 2021-04-27 ManuscriptTe.Rmd
Update format and references

---
title: "Manuscript Te example"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette: default
---

>>>>

---
title: "Manuscript Te example"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
  rmarkdown::html_vignette: default
---

### 2021-04-27 Manuscript_full.Rmd
Update format and references

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi,..."
date: "February 4, 2020"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

@@

@@

Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground. However, most organisms live near the ground and acute exposure to solar radiation can substantially elevate their body temperatures. We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR). The package includes microclimate models to vertically scale weather station data to organismal heights. The biophysical modelling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa. Additional functions model and temporally partition air and soil temperatures and solar radiation. The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules.

## Background
Many responses of organisms and ecosystems to climate change are inconsistent with current predictive models [@maguire2015modeling]. Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic]. Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m). However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [@kearney2009potential; @sinclair2016can; @sunday2014thermal].  Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions. Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life]. We aim to help remedy these shortcomings by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of the tools has not kept pace with research on climate change responses. Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].  Increasing dissemination of computational tools is facilitating adoption. The TrenchR package is intended to complement the NicheMapR package, which includes sophisticated microclimate [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models. Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.  The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.  Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2019thermimage]. Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals. We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: [@muir2019tealeaves]; plantecophys [@duursma2015plantecophys]].

## Methods and features
The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users. Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github. Issues and feature requests can be contributed in Github (https://github.com/trenchproject/TrenchR/). The package is maintained and checked using Travis CI as a tool for continuous integration.

Functions were adapted from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles. The source reference is included in each function’s header. We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available. Many functions additionally include comments that describe potential parameterizations. The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR [Table 8.2XXXX].

### Components
TrenchR functions are organized into the following categories:
**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.

@@

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss. Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.

@@

### Vignettes and introductory modules
We introduce the function in several vignettes.  A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* tutorial provides resources for estimating the environmental conditions experienced by organisms.  This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles. Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*. Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we have updated a series of educational modules entitled “Physical processes in ecosystems”.  We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical]. The modules provide less detailed coverage of many of the topics contained in the books. The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling. Thermodynamics and energy budget modelling are then introduced with detailed examples. Modules address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. We have revised the modules (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits. The modules are provided in html form and available for download as a pdf at the bookdown server (https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/). The R markdown files are available in Github at https://github.com/trenchproject/TrenchRmodules.

## Examples
We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial. We first estimate the energy budget with component functions and we then use an integrated biophysical model to estimate operative environmental temperatures. We start by generating environmental inputs. See Appendix S1 for another example examining a time series of microcliamte data and applying and testing the biophysical models.

@@

Let us first prepare the environmental data for analysis. We will need to estimate hourly air and soil temperatures and radiation. We start by estimating the day of year and the timing of sunrise and sunset:
```{r}
library(TrenchR)

Tb_CampbellNorman=function(T_a, T_g, S, alpha_L=0.96, epsilon=0.96, c_p=29.3, D, V){

@@

```{r}
#set up input data as variables
lat=35.69; lon= -105.944; elev=2121 #latitude and longitude (degrees) and elevation (m)
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 #minimum and maximum of air and soil temperatures, respectively (C)
V=1 #wind speed (m/s)
#assumptions
tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
Tb0=25 #initial assumption of body temperature (C)

@@

Although measured solar radiation is preferable if available, we can estimate hourly solar radiation by discounting incoming solar radiation as it moves through the atmosphere as follows:
```{r}
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) #estimate zenith angle (degrees)
psi_rad= degree_to_radian(psi_deg) #convert to radians
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho) #estimate radiation
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]; Sdif= Srad[2,]; Sref= Srad[3,]
```

We can then calculate hourly air and soil temperatures based on daily minimum and maximum temperatures. We select the sine-exponential model for air temperature and the sine model for surface temperature:
```{r}
Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) #air temperature (C)
Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) #soil temperature (C)
```

@@

### Energy balance
We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W). We will estimate each term on the right side of the equation in turn. Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical]. We will initially assume $T_b=T_a$ to illustrate the calculations before solving for $T_b$ givne the environmental conditions.

@@

We calculate the hourly solar radiation absorbed (W) as follows:
```{r}
Qabs= rep(NA, 24)
for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
```

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We assume the surface emissivity of lizards, $epsilon_s=0.965$ [@barlett1967energy].

@@

We estimate convective heat exchange between the animal and surrounding air using the following relationship:
$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$
where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions. We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. We note we currently estimate no convection since we are assuming $Tb=Ta$ for illustration purposes:
```{r}
Qconv= rep(NA, 24)
for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta_liz[hr]+273.15, T_b= Ta_liz[hr]+273.15, H=H_L, A=A, proportion=0.67, ef=1.3)
```

We estimate conductive heat flow (W) from the lizard to the surface assuming conductance through the animal tissue is the rate limiting step as follows:
$$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
We implement the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thickness $d$ of $10^{-6}$:
```{r}
Qcond= rep(NA, 24)
for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15, T_b=Ta_liz[hr]+273.15, d=10^-6, K=K, A=A, proportion=0.33)
```

We assume, as in generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible:
```{r}
Qmet= 0; Qevap= 0
```

The full heat budget can be calculated as follows [@gates1980biophysical]: 
```{r}
Qnet= Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
```
We constructed the energy balance assuming $T_b=T_a$ for illustration. Thus, the energy balance is dominated by heat gain from contact with the warmer surface. We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:

@@

We also implement a similar but simplified energy balance [@campbell2000introduction]. The energy balance omits conduction with the ground:
```{r}
Te2= rep(NA, 24)

@@

We additionally estimate $T_e$ using a specialized function for lizards [@buckley2008linking]:
```{r}
Te3= rep(NA, 24)
for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta_liz[hr], T_g=Ts[hr], u=V_liz, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
```

### Plot
We next plot the diurnal trends in environmental data and predicted body temperatures.

```{r}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) #Gates
points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") #Campbell and Norman
points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") #Lizard, Buckley 2008
points(1:24, Ta, type="l", col="#440154FF")
points(1:24, Ta_liz, type="l", col="#3B528BFF")
points(1:24, Ts, type="l", col="#21908CFF")

@@

The microclimate models indicate that air temperatures at lizard height are similar to surface temperatures (Figure 1). The biophysical models indicate that solar radiation will elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 1). The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model before selection.

## Summary
TrenchR is intended to promote understanding of how organisms interact with their environment and the consequences for physiology, energetics, behavior, and demography. Our example implementation highlights the importance of considering organismal body temperatures, rather than air temperatures, when examining thermal stress and other responses to environmental variability and change. TrenchR currently focuses on heat balances but may be expanded to include water balances. Simple functions can be combined as needed to produce comprehensive and transparent models for biophysical ecology and evolution. The resultant models are likely to be sufficiently detailed for many applications, but users are referred to NicheMapR for more detailed biophysical models [@kearney2017nichemapr; @kearney2020nichemapr]. Making classic biophysical ecology techniques more accessible will allow researchers to take advantage of rapidly accumulating data on environmental conditions and organismal traits to understand and predict ecological and evolutionary responses to variable and changing environments.

>>>>

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abby Meyer, Yutaro Sakairi,..."
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

-   Within the word count (6000-7000 words for Standard Articles, 3000
    words for Applications and Practical Tools)

-   Continuous line and page numbering throughout

Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^,...

1.  Department of Biology, University of Washington, Seattle, WA
    98195-1800, USA
2.

Corresponding Author: Lauren Buckley,
[lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}

Running head: TrenchR microclimate and biophysical models

Type: applications

@@

1.  Much understanding of climate change responses relies on the
    assumption that organismal body temperatures are equal to temporally
    averaged air temperatures high above the ground. However, most
    organisms live near the ground and acute exposure to solar radiation
    can substantially elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses
    (<https://github.com/trenchproject/TrenchR>). The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation.
3.  [Main results]
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting and includes tutorials and well as a series of
    educational modules.

**Keywords**: biophysical, climate change, ecotherm, energy budget,
mechanistic models, microclimate, scaling, temperature

## Introduction

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperate at weather station height (usually 2m). However, air
temperatures near the ground where most organisms reside can be
considerably warmer and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
@kearney2009]. Acute thermal stress events are often more relevant to
the physiology and energy balance of organisms than daily, monthly, or
annually aggregated environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package as a component of the TrEnCh project
(<https://trenchproject.github.io>), which builds computational and
visualization tools to Translate Environmental Change into organismal
responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
tools has not kept pace with research on climate change responses.
Adoptions of biophysical approaches are expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr]. Increasing dissemination of computational tools
is facilitating adoption. The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020] models. Many of
the approaches are similar between the packages, but TrenchR focuses on
simple functions to aid the understanding and transparency of
biophysical approaches. The TrenchR functions allow for modular use and
easy adaptation to a variety of organismal and research questions. Some
energy budget calculation components of biophysical models are also
available in the ThermImage package [@tattersall2017]. Although many of
the TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We have minimized inclusion of plant
specific biophysical models since they are the focus of several R
packages [tealeaves: [@muir2019tea]; plantecophys:
[@duursma2015plantecophys]].

## Methods and features

The TrenchR package aims to promote transparency and reproducibility and
we welcome contributions and corrections from users. Our package is
built using the devtools methodology
(<https://github.com/hadley/devtools>) in Github. Issues and feature
requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR/>). The package is maintained
and checked using Travis CI as a tool for continuous integration.

Functions were adapted from biophysical ecology textbooks
[@campbell2000introduction; @gates1980biophysical] and research
articles. The source reference is included in each function's header. We
have incorporated default parameters (e.g., organism emissivity or heat
conduction rates) when general values are available. Many functions
additionally include comments that describe potential parameterizations.
The biophysical ecology texts [@campbell2000introduction;
@gates1980biophysical] include many tables with parameter values and we
provide several tables in TrenchR [Table 8.2XXXX].

### Components

TrenchR functions are organized into the following categories:

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles), facilitate modeling diurnal temperature variation,
and estimate biologically relevant aggregate metrics such as degree days
available for development.

@@

**General biophysical functions** provide generalized models of heat
exchanges between organisms and their environment so that users can
build custom biophysical models. The functions allow implementing an
energy balance of the following components of heat exchange between
organisms and their environment: *radiative heat exchanges* of solar and
thermal radiation; *convective heat exchanges* between organisms and
their surrounding fluid (air or water) driven by fluid flow; *conductive
heat exchanges* between organisms and the ground due to physical
contact; and *evaporative heat exchanges* associated with organisms'
evaporative water loss. Additional functions aggregate these forms of
heat exchange into energy balances and to use the energy balances to
predict body temperatures.

@@

### Vignettes and introductory modules

We introduce the function in several vignettes. A good place to start is
the *Allometry and conversions* tutorial, which provides tools for
preparing data such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* tutorial
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and windspeed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we
have updated a series of educational modules entitled "Physical
processes in ecosystems". We intend the modules to provide a more
contained and accessible introduction to microclimate and biophysical
modelling than that included in classic textbooks
[@campbell2000introduction; @gates1980biophysical]. The modules provide
less detailed coverage of many of the topics contained in the books. The
series of 14 modules starts with an overview of the calculus and physics
principles underlying the modelling. Thermodynamics and energy budget
modelling are then introduced with detailed examples. Modules address
the climate space concept, operative temperatures, biophysical models
for leaves and sheep, and heat flow in soils. We have revised the
modules (which originated from an NSF training grant in 1979 and lacked
broad distribution) to include R code and utilization of TrenchR
functions. We expanded the original series of tutorials to include
introductions by Michael Kearney to the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
modules are provided in html form and available for download as a pdf at
the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github at
<https://github.com/trenchproject/TrenchRmodules>.

## Examples

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
incorporated in the *Using energy balances to estimate body
temperatures* tutorial. We first estimate the energy budget with
component functions and we then use an integrated biophysical model to
estimate operative environmental temperatures. We start by generating
environmental inputs. See Appendix S1 for another example examining a
time series of microclimate data and applying and testing the
biophysical models.

@@

Let us first prepare the environmental data for analysis. We will need
to estimate hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

```{r, echo=FALSE}
library(TrenchR)

#to be REMOVED
Tb_CampbellNorman=function(T_a, T_g, S, alpha_L=0.96, epsilon=0.96, c_p=29.3, D, V){

@@

```{r}
#set up input data as variables
#latitude and longitude (degrees) and elevation (m)
lat=35.69; lon= -105.944; elev=2121 
#minimum and maximum of air and soil temperatures, respectively (C)
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 
#wind speed (m/s)
V=1 
#assumptions
tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
Tb0=25 #initial assumption of body temperature (C)

@@

Although measured solar radiation is preferable if available, we can
estimate hourly solar radiation by discounting incoming solar radiation
as it moves through the atmosphere as follows:

```{r}
#estimate zenith angle (degrees)
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) 
#convert to radians
psi_rad= degree_to_radian(psi_deg) 
#estimate radiation
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]; Sdif= Srad[2,]; Sref= Srad[3,]
```

We can then calculate hourly air and soil temperatures based on daily
minimum and maximum temperatures. We select the sine-exponential model
for air temperature and the sine model for surface temperature:

```{r}
#air temperature (C)
Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) 
#soil temperature (C)
Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) 
```

@@

### Energy balance

We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10g
*Sceloporus* lizard with solar absorptivity $a=0.9$
[@gates1980biophysical]. We will initially assume $T_b=T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

@@

We calculate the hourly solar radiation absorbed (W) as follows:

```{r}
Qabs= rep(NA, 24)
for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
```

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
assume the surface emissivity of lizards, $epsilon_s=0.965$
[@barlett1967].

@@

We estimate convective heat exchange between the animal and surrounding
air using the following relationship:
$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
enhancement factor, $ef$, multiplier can be incorporated to account for
increases in heat exchange resulting from air turbulence in field
conditions. We implement the function in R assuming that 2/3 of the
lizard's surface area is exchanging heat through convection. We note we
currently estimate no convection since we are assuming $Tb=Ta$ for
illustration purposes:

```{r}
Qconv= rep(NA, 24)
for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta_liz[hr]+273.15, T_b= Ta_liz[hr]+273.15, H=H_L, A=A, proportion=0.67, ef=1.3)
```

We estimate conductive heat flow (W) from the lizard to the surface
assuming conductance through the animal tissue is the rate limiting step
as follows: $$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
We implement the estimate in R assuming that 1/3 of the lizard surface
is in contact with the ground and a skin thickness $d$ of $10^{-6}$:

```{r}
Qcond= rep(NA, 24)
for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15, T_b=Ta_liz[hr]+273.15, d=10^-6, K=K, A=A, proportion=0.33)
```

We assume, as in generally done for lizards, that heat exchange
associated with metabolism and evaporation is negligible:

```{r}
Qmet= 0; Qevap= 0
```

The full heat budget can be calculated as follows
[@gates1980biophysical]:

```{r}
Qnet= Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
```

We constructed the energy balance assuming $T_b=T_a$ for illustration.
Thus, the energy balance is dominated by heat gain from contact with the
warmer surface. We now use a function based on the energy balance above
to estimate body temperature given the environmental conditions:

@@

We also implement a similar but simplified energy balance
[@campbell2000introduction]. The energy balance omits conduction with
the ground:

```{r}
Te2= rep(NA, 24)

@@

We additionally estimate $T_e$ using a specialized function for lizards
[@buckley2008link]:

```{r}
Te3= rep(NA, 24)
for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta_liz[hr], T_g=Ts[hr], u=V_liz, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
```

### Plot

We next plot the diurnal trends in environmental data and predicted body
temperatures.

```{r, echo=FALSE}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) #Gates
points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") #Campbell and Norman
points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") #Lizard, Buckley 2008
points(1:24, Ta, type="l", col="#440154FF")
points(1:24, Ta_liz, type="l", col="#3B528BFF")
points(1:24, Ts, type="l", col="#21908CFF")

@@

The microclimate models indicate that air temperatures at lizard height
are similar to surface temperatures (Figure 1). The biophysical models
indicate that solar radiation will elevate lizard body temperatures far
above air temperatures and that the lizard will face thermal stress if
it is unable to seek shade (Figure 1). The three biophysical models
predict different body temperatures during peak period of solar
radiation because they model interactions with the ground differently
and users are encouraged to review the details of each biophysical model
before selection. Differences in estimated body temperatures are
accentuated by the high level of solar radiation.

## Conclusions

TrenchR is intended to promote understanding of how organisms interact
with their environment and the consequences for physiology, energetics,
behavior, and demography. Our example implementation highlights the
importance of considering organismal body temperatures, rather than air
temperatures, when examining thermal stress and other responses to
environmental variability and change. TrenchR currently focuses on heat
balances but may be expanded to include water balances. Simple functions
can be combined as needed to produce comprehensive and transparent
models for biophysical ecology and evolution. The resultant models are
likely to be sufficiently detailed for many applications, but users are
referred to NicheMapR for more detailed biophysical models
[@kearney2017nichemapr; @kearney2020nichemapr]. Making classic
biophysical ecology techniques more accessible will allow researchers to
take advantage of rapidly accumulating data on environmental conditions
and organismal traits to understand and predict ecological and
evolutionary responses to variable and changing environments.

## Acknowledgements

We thank Brian Helmuth and Eric Riddell for providing and checking
functions. We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
Porter, and Art Woods for assistance and input. We particularly thank
Michael Kearney for extensive input on TrenchR and its alignment with
NicheMapR and for contributing two educational modules.

[Check with Eric]

## Authors' contributions

LB conceived and designed the package and lead all aspect of the
packages creation and dissemination; BBO, IC, AJ, OL, AM, YS wrote and
edited functions and associated package resources; AJ configured the
package; AM, BBO, YS, and LB converted and edited the educational
modules. All authors contributed critically to the drafts and gave final
approval for publication.

## Data availability

All code is available at <https://github.com/trenchproject/TrenchR>.
